Best Practices in Latent Class Analysis using Free Open Source Software

### 2023-02-23 manuscript.Rmd
edits from to do list, MGV

Finally, LCA can be used to deal with data which violate certain assumptions.
For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughman_mixture_2007; @magnus_multidimensional_2022].
These models estimate one class to explain the "true zeros",
and another class with a different distribution to explain all other values, which can include zero.
Note that this LCA has different models for both classes,
and thus falls beyond the scope of this paper.
<!-- CJ: This has been said elsewhere: -->
<!-- In fact, LCA assumes the population -->
<!-- distribution is a non-normal mixture of $k$ normal distributions, and it -->
<!-- can recover the value of $k$, i.e. generate a $k$-class solution. -->
<!-- CJ: This example is not perfect, we need another example still.. -->
For instance, the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality [@alkema_maximizing_2022].
A mixture of two normal distributions with distinct variances described the observed distribution better.
<!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical. We encourage researchers to be agnostic about the **true** nature of the latent variable. And focus on the approach that fits the research question. -->
<!-- CJ: Why would we encourage researchers to be agnostic about the true...? -->

@@

Since SA is a global optimizer, it should be independent of starting values.
With this in mind, it is efficient to start estimation from a "reasonable solution", instead of a random starting point.
For example, if we assume that the different classes are likely to have different mean values on the indicators, 
then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids [see also @biernacki_choosing_2003].
By default, `tidySEM` derives starting values using K-means clustering [@likas_global_2003],
computing the mixture model starting values by treating the K-means solution as multi-group model.
Next, SA is used to find the global optimum solution.
Finally, SA is followed up with a short run of the ML algorithm,
as ML inherently produces an asymptotic
covariance matrix for the parameters that can be used to compute
standard errors.
Note that these defaults can be manually overridden.

@@

Due to the lack of absolute fit indices for LCA, the assessment of relative fit
relies on information criteria.
Information criteria (ICs) are computed from the $-2LL$,
which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
In addition to capturing model fit, ICs add a penalty for model complexity thus
preferring more model fit and parsimony.
The lower the value of an information criterion, the better the overall fit of the
model.
The *Akaike Information Criterion* (*AIC*) is the original information criterion.
It is computed as $-2LL+2p$, where $p$ is the number of parameters.
The *Bayesian Information Criterion* (*BIC*) multiplies the penalty for complexity with a function of the sample size, $p \ln(n)$,
such that complex models are penalized more in smaller samples.
Finally,
the so-called *sample size adjusted BIC* (*saBIC*) implements a different ad-hoc penalty for sample size,
$\ln(\frac{n + 2}{24})$. Although simulation studies have showen that the saBIC performs well for class enumeration [@tein_statistical_2013; @chen_efficacy_2017],
a limitation is that its penalty is not based on theory. BIC also performs well and is theoretically substantiated.
As such, BIC may be the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
In `tidySEM`, all three ICs are available as a part of the output of `table_fit()`.

@@

Second, most likely class membership based on the highest posterior class probability, $F_M$, is obtained by running `class_prob(res, "sum.mostlikely")`.
This is the frequency table of the variable $M$.
Unlike $F_P$, $F_M$ ignores classification error.
If every observation is classified with perfect accuracy,
$F_M$ and $F_P$ are identical.
Part of the output of `table_fit()` for LCA analyses
are the minimum and maximum percentage of the sample assigned to a particular class, `n_min` and `n_max`, derived from $F_M$.
This is reported by default because the smallest class proportion, `n_min`, is of particular interest.
When the size of the smallest class is not much larger than the number of parameters per class,
proceed with caution:
The model might not be locally identified in that class [@depaoli_mixture_2013].
Existing rules of thumb have focused on smallest class size in terms of percentage of the total sample size.
This is not particularly meaningful,
however, as the primary concern is that small classes may comprise too few observations to accurately estimate class parameters.
Thus, a better rule of thumb might be to ascertain that the smallest class has many more observations than the average number of parameters per class.
<!-- DA: what is many more? When making a rule a thumb, you have to eyeball a quantity and recommend it for users without statistical intuition-->
Another good practice is to ensure that the smallest class has a unique and meaningful substantive interpretation.
If the smallest class does not comprise at least a few observations per parameter, or is redundant with another larger class, a simpler solution is preferable.

To assess classification accuracy, we turn to the third diagnostic table:
the average posterior probabilities by most likely class membership, $P_{p_{(M = m)}}$
which can be obtained using `class_prob(res, type = "avg.mostlikely")`.
For participants assigned to one particular class (rows) according to $M$,
this table gives the mean probability of being assigned to any class (columns).
This table is obtained by calculating the column means of $P$ for the 
$k$ subsets of observations with most likely class of $M = 1 \ldots k$.
The diagonal represents the average predictive probability of the subjects related 
to their most likely class, indicating the certainty of class assignment.
Some guidelines have being suggested for this diagonal, 
such as its values having to be higher than 0.7 [@masyn_latent_2013]. 
We, however, advise caution as these guidelines cannot be generalized to all situations,
and should not be used as strict cutoffs communication the classification quality.

@@

Classification diagnostics should be reported,
as they may determine whether models are excluded from consideration.
Of particular interest is the diagonal of either the average posterior probabilities by most likely class membership,
or of the most likely class membership by true class membership.
The former indicates the average probability that individuals belong to the class they were assigned to,
and the latter indicates the probability that individuals are assigned to a class if they actually belong to that class.
By default, `table_fit()` reports the latter as `prob_min` and `prob_max`.
The entropy is also of interest as an overall measure of class separation.
Finally, the range of proportions of cases assigned to each class is important;
the minimum proportion of cases per class can be used to diagnose classes that are too small to be locally identified or practically relevant [see @depaoli_mixture_2013].
By default, `table_fit()` reports the minimum and maximum proportion of cases,
based on most likely class membership.
Fifth, the LMR or BLRT can be reported if it is a part of the class enumeration strategy, using `lr_lmr()` and `BLRT()`.
By default, `table_fit()` reports the VLM for subsequent models in the table.

>>>>

Finally, LCA can be used to deal with data which violate certain assumptions.
For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughman_mixture_2007; @magnus_multidimensional_2022].
These models estimate one class to explain the "true zeros",
and another class with a different distribution to explain all other values, which can include zero.
Note that this LCA has different models for both classes,
and thus falls beyond the scope of this paper.
<!-- CJ: This has been said elsewhere: -->
<!-- In fact, LCA assumes the population -->
<!-- distribution is a non-normal mixture of $k$ normal distributions, and it -->
<!-- can recover the value of $k$, i.e. generate a $k$-class solution. -->
<!-- CJ: This example is not perfect, we need another example still.. -->
For instance, a depression questionnaire PH-9 have found 2 distinct groups, ones with no symptoms, and one with symptoms. A mixture model can model the "presence" and "frequency" of symptoms as separate classes, instead of assuming that the sample follows a single continuous underlying distribution [@magnus_multidimensional_2022]. 
<!-- the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality [@alkema_maximizing_2022].
A mixture of two normal distributions with distinct variances described the observed distribution better. -->
<!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical. We encourage researchers to be agnostic about the **true** nature of the latent variable. And focus on the approach that fits the research question. -->
<!-- CJ: Why would we encourage researchers to be agnostic about the true...? -->

@@

Since SA is a global optimizer, it should be independent of starting values.
With this in mind, it is efficient to start estimation from a "reasonable solution", instead of a random starting point.
For example, if we assume that the different classes are likely to have different mean values on the indicators, 
then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids [see also @biernacki_choosing_2003].
By default, `tidySEM` derives starting values using K-means clustering [@likas_global_2003; @Forgy1965ClusterAO; @Hartigan_kmeans],
computing the mixture model starting values by treating the K-means solution as multi-group model.
Next, SA is used to find the global optimum solution.
Finally, SA is followed up with a short run of the ML algorithm,
as ML inherently produces an asymptotic
covariance matrix for the parameters that can be used to compute
standard errors.
Note that these defaults can be manually overridden.

@@

Due to the lack of absolute fit indices for LCA, the assessment of relative fit
relies on information criteria.
Information criteria (ICs) are computed from the $-2LL$,
which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
In addition to capturing model fit, ICs add a penalty for model complexity thus
preferring more model fit and parsimony.
The lower the value of an information criterion, the better the overall fit of the
model.
The *Akaike Information Criterion* (*AIC*) is the original information criterion.
It is computed as $-2LL+2p$, where $p$ is the number of parameters.
The *Bayesian Information Criterion* (*BIC*) multiplies the penalty for complexity with a function of the sample size, $p \ln(n)$,
such that complex models are penalized more in smaller samples.
Finally,
the so-called *sample size adjusted BIC* (*saBIC*) implements a different ad-hoc penalty for sample size,
$\ln(\frac{n + 2}{24})$. Although simulation studies have shown that the saBIC performs well for class enumeration [@tein_statistical_2013; @chen_efficacy_2017],
a limitation is that its penalty is not based on theory, but on the Minimum Description Length (MDL) as the shortes set of information necessary to reproduce the data [@Rissanen_1985; @sclove_application_1987]. BIC also performs well and is theoretically substantiated.
As such, BIC may be the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
In `tidySEM`, all three ICs are available as a part of the output of `table_fit()`.

@@

Second, most likely class membership based on the highest posterior class probability, $F_M$, is obtained by running `class_prob(res, "sum.mostlikely")`.
This is the frequency table of the variable $M$.
Unlike $F_P$, $F_M$ ignores classification error.
If every observation is classified with perfect accuracy,
$F_M$ and $F_P$ are identical.
Part of the output of `table_fit()` for LCA analyses
are the minimum and maximum percentage of the sample assigned to a particular class, `n_min` and `n_max`, derived from $F_M$.
This is reported by default because the smallest class proportion, `n_min`, is of particular interest.
When the size of the smallest class is not much larger than the number of parameters per class,
proceed with caution:
The model might not be locally identified in that class [@depaoli_mixture_2013].
Existing rules of thumb have focused on smallest class size in terms of percentage of the total sample size.
This is not particularly meaningful,
however, as the primary concern is that small classes may comprise too few observations to accurately estimate class parameters.
Thus, a better rule of thumb might be to ascertain that the smallest class has at least as many observations as the average number of parameters per class.
<!-- DA: what is many more? When making a rule a thumb, you have to eyeball a quantity and recommend it for users without statistical intuition
MGV: here I changed it "at least as many", as a low bound recommendation-->
Another good practice is to ensure that the smallest class has a unique and meaningful substantive interpretation.
If the smallest class does not comprise at least a few observations per parameter, or is redundant with another larger class, a simpler solution is preferable.

To assess classification accuracy, we turn to the third diagnostic table:
the average posterior probabilities by most likely class membership, $P_{p_{(M = m)}}$
which can be obtained using `class_prob(res, type = "avg.mostlikely")`.
For participants assigned to one particular class (rows) according to $M$ (most likely class based on predicted probabilities),
this table gives the mean probability of being assigned to each class (columns).
This table is obtained by calculating the column means of $P$ for the 
$k$ subsets of observations with most likely class of $M = 1 \ldots k$.
The diagonal represents the average predictive probability of the subjects related 
to their most likely class, indicating the certainty of class assignment.
Some guidelines have being suggested for this diagonal, 
such as its values having to be higher than 0.7 [@masyn_latent_2013]. 
We, however, advise caution as these guidelines cannot be generalized to all situations,
and should not be used as strict cutoffs communication the classification quality.

@@

Classification diagnostics should be reported,
as they may determine whether models are excluded from consideration.
Of particular interest is the diagonal of either the average posterior probabilities by most likely class membership,
or of the most likely class membership by true class membership.
The former indicates the average probability that individuals belong to the class they were assigned to,
and the latter indicates the probability that individuals are assigned to a class if they actually belong to that class.
By default, `table_fit()` reports the latter as `prob_min` and `prob_max`.
The entropy is also of interest as an overall measure of class separation.
Finally, the range of proportions of cases assigned to each class is important;
the minimum proportion of cases per class can be used to diagnose classes that are too small to be locally identified or practically relevant [see @depaoli_mixture_2013].
By default, `table_fit()` reports the minimum and maximum proportion of cases,
based on most likely class membership.
Fifth, the LMR or BLRT can be reported if it is a part of the class enumeration strategy, using `lr_lmr()` and `BLRT()`.
By default, `table_fit()` reports the LMR (also called Vuong’s robust likelihood-ratio test, VLMR) for subsequent models in the table.
